---
title: "Heatmap RNAseq"
output: pdf_document
date: "2023-11-04"
---
This script generates heatmaps from RNA sequencing (RNAseq) data, applying several data transformations and customizations. Here’s an overview of the main steps:

1. Load Libraries
2. Data Import and Preparation: RNAseq data is read from a CSV file, and transformations are applied: Log2 and Z-score Transformation. 
3. Heatmap Generation and Saving as PDF and PNG files with different dimensions to accommodate various presentation needs.
4. Troubleshooting: Commands are included to detect and locate any NA (missing) values, addressing common data issues that may prevent heatmap generation.

Based on the following scripts online: 
https://davetang.github.io/muse/pheatmap.html
https://bioinformatics.ccr.cancer.gov/docs/data-visualization-with-r/Lesson5_intro_to_ggplot/

Load necessary libraries
```{r}
library(pheatmap) 
library(tidyverse) 
library(ggplotify) 
library(heatmaply) 
```

Read the data table
```{r}
mat<-read.csv("/Users/jeffreyreina/Documents/Salk/RNAseq MDA-MB-231 results/03.Result_X202SC23073852-Z01-F001_Homo_sapiens/Analysis without Rescues/heatmap_replicates.csv",header=TRUE,row.names=1)
```

Transform data to log2 scale
```{r}
mattransformed<-log2(mat+1) #+1 to avoid errors with 0 values
```

Transform data to z-scores
```{r}
cal_z_score <- function(x){
  (x - mean(x)) / sd(x)
}

matzscore <- t(apply(mattransformed, 1, cal_z_score))
```

Define sample groups for annotation
```{r}
dfh<-data.frame(sample=as.character(colnames(mat)),group="Treatment")%>%
                column_to_rownames("sample")
dfh$group<-ifelse(rownames(dfh) %in% c("KO1","KO2","KO3"),
                "KO","WT")
dfh
```


Choose color palette for heatmap
```{r}
my_palette <- colorRampPalette(c("green", "black", "red"))(n = 9999)
```

Save heatmap as a PDF
```{r}
pdf("Heatmap2.pdf",width=4.4,height=6.3)

pheatmap(matzscore, color = my_palette, scale = "row", show_rownames = FALSE, annotation_col = dfh, annotation_colors=list(group=c(KO="blue",WT="orange")), treeheight_row = 10)
dev.off()
```

Save heatmap as a PNG
```{r}
png("pheatmap_5.png", res=300, width=7, height=4.5, unit="in")

pheatmap(matzscore, color = my_palette, scale = "row", show_rownames = FALSE, annotation_col = dfh, annotation_colors=list(group=c(KO="blue",WT="red")), treeheight_row = 10)
dev.off()
```

Save heatmap with exchanged width and height in PNG
```{r}
png("pheatmap_2.png", res=300, width=4.5, height=7, unit="in")

pheatmap(matzscore, color = my_palette, scale = "row", show_rownames = FALSE, annotation_col = dfh, annotation_colors=list(group=c(KO="blue",WT="red")),  treeheight_row = 10)
dev.off()
```

Save heatmap as an R object
```{r}
hm_ph <-pheatmap(matzscore, color = my_palette, scale = "row", show_rownames = FALSE, annotation_col = dfh, annotation_colors=list(group=c(KO="blue",WT="orange")),  treeheight_row = 10)
```

```

