---
title: "Volcano Plot for Proximity Labeling Proteomics"
output: pdf_document
date: "2023-11-05"
---

This R script generates a volcano plot to visualize differentially regulated proteins on a proximity-labeling proteomics (e.g TurboID) dataset, highlighting proteins based on significance (p-value) and regulation (up- or down-regulated).

Based on the following script: 
https://github.com/kevinblighe/EnhancedVolcano

Install necessary packages if not already installed

Install packages
```{r}
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("EnhancedVolcano")
```

Load data from a CSV file
```{r}
mat<-read.csv("/Users/jeffreyreina/Documents/Salk/TurboID analysis/MDATurboUniprot.csv",header=TRUE,row.names=1)
```

Ensure the log2 fold change column is numeric
```{r}
mat$log2fc<-as.numeric(mat$log2FC)
```

Load EnhancedVolcano package for plotting
```{r}
library(EnhancedVolcano)
```

Generate a basic volcano plot
```{r}
EnhancedVolcano(mat,
                lab = rownames(mat),
                x = "log2fc",
                y = "Pval") 
```

Check the first few rows of the matrix
```{r}
head(mat)  
```

Add a significance column based on p-value threshold
```{r}
mat$significant<-ifelse (mat$Pval <= 0.05, "true", "false")
```

Calculate -log10 of the p-value for plotting
```{r}
mat$log10pval<--log10(mat$Pval)
```

Identify upregulated and downregulated genes
```{r}
mat$updown<-ifelse (mat$FC <= 1, "down", "up")
```


Create a volcano plot with significant proteins labeled and colored

```{r}

ggplot(data=mat, aes(x = log2FC, y = log10pval)) + 
  geom_point(data = subset(mat, significant %in% c("true")), size = 2,  
              alpha = 0.5,  aes(color = updown), shape=19) +
  scale_color_manual(values = c("#2c538f", "firebrick")) + geom_point(data = subset(mat, significant %in% c("false")), size = 2,  
                alpha = 0.5,  colour = "#3a3d3a") +
  theme_bw() + 
  geom_text_repel(data = subset(mat, significant %in% c("true")), aes(label = Gene_Name))
```


Save the plot as a PDF
```{r}
pdf("volcanoTurboID.pdf", width=6, height=4.5)

ggplot(data=mat, aes(x = log2FC, y = log10pval)) + 
  geom_point(data = subset(mat, significant %in% c("true")), size = 2,  
              alpha = 0.5,  aes(color = updown), shape=19) +
  scale_color_manual(values = c("#2c538f", "firebrick")) + geom_point(data = subset(mat, significant %in% c("false")), size = 2,  
                alpha = 0.5,  colour = "#3a3d3a") +
  theme_bw() + 
  geom_text_repel(data = subset(mat, significant %in% c("true")), aes(label = Gene_Name))
```
